D3064R0: Bringing the Laws of Physics, Computer-System Constraints, and Compilers to Bear on the OOTA Problem

### 2024-02-13 ootavacuous.tex
February 13 review of reworked material
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

For accesses to partially volatile objects, we remove some of the
requirements and add a new one.
We leave out the last two items; compilers are allowed to merge or
omit accesses to partially volatile objects.
And the requirement for order preservation is weakened; it applies
only to pairs of accesses to the same object.
Accesses to different objects may be reordered relative to each other.
The new requirement is that the map must preserve the rf relation.
% Is this rf?  Or specifically rfe?
That is, if load $R$ in $A$ reads from store $W$ and $R$ is not
omitted then the instruction it maps to in $H$ must read from the
instruction that $W$ maps to,
and consequently $W$ must not be omitted either.

@@

Appendix~\ref{app:Informal Definition of Semantic Dependency}
outlines the informal informal definition of ``semantic dependency''
used to evaluate the litmus tests in this section,
Appendix~\ref{app:Litmus Tests from “Causality Test Cases"}
evaluates the obligatory litmus tests from the JMM Causality Test
Cases, and
Appendix~\ref{app:Additional Litmus Tests}
evaluates some additional litmus tests that affected our thinking or
otherwise gave us pause.

@@

@@

>>>>

For accesses to partially volatile objects, we remove some of the
requirements and add a new one.
We leave out the last two items; compilers are allowed to merge or
omit accesses to partially volatile objects.
And the requirement for order preservation is weakened; it applies
only to pairs of accesses to the same object.
Accesses to different objects may be reordered relative to each other.
The new requirement is that the map must preserve the rf relation.
That is, if load $R$ in $A$ reads from store $W$ and $R$ is not
omitted then the instruction it maps to in $H$ must read from the
instruction that $W$ maps to,
and consequently $W$ must not be omitted either.

@@

Appendix~\ref{app:Litmus Tests from “Causality Test Cases"}
evaluates the obligatory litmus tests from the JMM Causality Test
Cases, and
Appendix~\ref{app:Additional Litmus Tests}
evaluates some additional litmus tests that affected our thinking or
otherwise gave us pause.

@@

\subsubsection{Semantic Dependencies and \co{volatile}}
\label{app:Semantic Dependencies and volatile}

The Linux kernel uses volatile accesses to constrain the compiler,
and it is worth looking at the example shown in
Listing~\ref{lst:Non-Volatile Accesses and Dependencies}
to see how this works.

\begin{listing}
\begin{verbatim}
 1 atomic<int> x, y;
 2
 3 void thread1() {
 4   int r1 = x.load(memory_order::relaxed);
 5   y.store(r1, memory_order::relaxed);
 6 }
\end{verbatim}
\caption{Non-Volatile Accesses and Dependencies}
\label{lst:Non-Volatile Accesses and Dependencies}
\end{listing}

Because both \co{x} and \co{y} are non-volatile, the compiler is
free to assume that the value loaded by line~4 will be some value
stored to \co{x} or its initial value.
Because there are no stores to \co{x}, the only remaining possibility
is the default-initialized value of zero.
Therefore, the C++ implementation is within its rights to substitute
the constant zero for \co{r1} on line~5, eliminating the semantic
dependency that would otherwise extend from line~4 to line~5.
This in turn enables the implementation to reorder line~5 to
precede line~4.

\begin{listing}
\begin{verbatim}
 1 atomic<volatile int> x, y;
 2
 3 void thread1() {
 4   int r1 = x.load(memory_order::relaxed);
 5   y.store(r1, memory_order::relaxed);
 6 }
\end{verbatim}
\caption{Volatile Accesses and Dependencies}
\label{lst:Volatile Accesses and Dependencies}
\end{listing}

One way to preserve this dependency is use of \co{volatile}, as shown in
Listing~\ref{lst:Volatile Accesses and Dependencies}.
Now the implementation is forbidden from assuming that the value loaded
by line~4 has any relation to any stores to or initialization of \co{x},
even if \co{x} is placed in normal memory.

In this case, the \co{volatile} keyword alerts the implementation to
the possibility of interfering accesses to \co{x} from outside the
program, for example, due to \co{x} being:

\begin{itemize}
\item	Allocated in an MMIO region of the address space.
\item	Modified by an unknown-to-the-implementation thread
	or signal handler.
\item	Subject to I/O-device DMA operations.
\item	Modified by debugger commands.
\item	Modified by accesses to facilities such as \path{/dev/mem}.
\item	Modified by as-yet-unwritten dynamically linked libraries.
\item	Modified by some other mechanism of the reader's choosing.
\end{itemize}

These possibilities force the implementation to preserve the semantic
dependency from line~4 to line~5.
They can also prevent the formation of OOTA cycles local to the
implementation because such inteference means that the volatile loads
might be reading from something that the implementation knows nothing
about.

The developer is free to arrange for \co{x} to be free of any
interference, thus obtaining predictable behavior along with the
preserved semantic dependency.
However, this example invalidates the rough definition of semantic
dependency from
Section~\ref{sec:Simple OOTA Cycle}
because the value loaded from \co{x} will always be the initial value
of zero, so that it is at best dubious to talk about any changes in
the value loaded.

One way forward is to assume that interference could happen when
determining which dependencies are semantic, and then do further analysis
using that determination.

@@

But first, Appendix~\ref{app:Informal Definition of Semantic Dependency}
outlines an informal informal definition of ``semantic dependency''
that can be used in \co{cbmc} test cases.
